
```{r}

```

# Results
Let’s look at the data, Image 1 best illustrates this answer. The rate of death with no
heart transplant drops significantly over time. The number of patients who did not receive a
transplant is 32, with only 2 surviving past 500 days. Patients who did receive the transplant are
67, with more than 22 surviving past 500 days.
In Image 2 categories patients under 40, between 40-50 and over 50 years old and a
hazard ratio of 2.08. From one category to the next this tells us you are twice as likely to die.
Not accounting for if you get the transplant or not.
First model excluded wait time as wait time is only available for patients who received
transplant, and not all patients in the study had the opportunity to receive a heart transplant. The
C-statistic =0.747 (see Table 1).

## Code


Second model generated included wait time to see if wait time is significant, only looking at
records that received the transplant, not the control who did not receive transplant. The result
C-statistic =0.683 (see Table 2).

## Code
```{r}
#generate model with wait time to see if wait time is significant, only looks at records that
received the transplant not the control who did not receive transplant
treated.mod <- coxph(Surv(survtime, survived)~ acceptyear + age + prior + wait)
```
Third model without an accepted year was generated and then compared using an ANOVA
table (see Table 4). The p-value = 0.1656 which is greater than 0.05 and we can drop the
accepted year as a variable.

## Code
```{r}
#generate model without accepted year
all2.mod <- coxph(Surv(survtime, survived)~ age + prior + transplant)
```

The C-statistic for the model without an accepted year is equal to 0.739 (see Table 3). Which
shows that this model has a higher predictability capabilities compared to the model including
the accepted year with a C-statistic = 0.683.

## Code
```{r}
#examine model
summary(all2.mod)
#compare model that includes acceptedyear and that does not
#high p value, there is not a statistically significant difference between the two models, can
drop the variable
anova(all.mod, all2.mod, test="LRT")

```


The C-statistic for the model without an accepted year is equal to 0.739 (see Table 3). Which
shows that this model has a higher predictability capabilities compared to the model including
the accepted year with a C-statistic = 0.683.