```{r}
library(survival)
#install.packages("survminer")
library(survminer)
```

```{r}
heart_transplant <- read.csv("./heart_transplant.csv")
#generate numeric columns for survived, prior, and transplant from existing charachter columns
heart_transplant$survived2 <- as.numeric(ifelse(heart_transplant$survived=="dead",1,0)) > 0
heart_transplant$prior2 <- as.numeric(ifelse(heart_transplant$prior=="yes",1,0))
heart_transplant$transplant2 <- as.numeric(ifelse(heart_transplant$transplant=="treatment",1,0))
all_strTrtmnt.mod <- coxph(Surv(survtime, survived2)~ age + prior2 + strata(transplant2), data=heart_transplant)
summary(all_strTrtmnt.mod)
```


# Results

Let's look at the data, Image 1 best illustrates this answer. The rate of death with no heart transplant drops significantly over time. The number of patients who did not receive a transplant is 32, with only 2 surviving past 500 days. Patients who did receive the transplant are 67, with more than 22 surviving past 500 days. In Image 2 categories patients under 40, between 40-50 and over 50 years old and a hazard ratio of 2.08. From one category to the next this tells us you are twice as likely to die. Not accounting for if you get the transplant or not. First model excluded wait time as wait time is only available for patients who received transplant, and not all patients in the study had the opportunity to receive a heart transplant. The C-statistic =0.747 (see Table 1).
## Code
#generate model without wait time #wait time is only available for those who received transplant 
```{r}
all.mod <- coxph(Surv(survtime, survived2)~ acceptyear + age + prior + transplant, data=heart_transplant)
```

Second model generated included wait time to see if wait time is significant, only looking at records that received the transplant, not the control who did not receive transplant. The result C-statistic =0.683 (see Table 2).

## Code

```{r}
treated.mod <- coxph(Surv(survtime, survived2)~ acceptyear + age + prior + wait, data=heart_transplant)
```

Third model without an accepted year was generated and then compared using an ANOVA table (see Table 4). The p-value = 0.1656 which is greater than 0.05 and we can drop the accepted year as a variable.

## Code

```{r}
#generate model without accepted year
all2.mod <- coxph(Surv(survtime, survived2)~ age + prior + transplant, data=heart_transplant)
```

The C-statistic for the model without an accepted year is equal to 0.739 (see Table 3). Which shows that this model has a higher predictability capabilities compared to the model including the accepted year with a C-statistic = 0.683.

## Code

```{r}
#examine model
summary(all2.mod)
#compare model that includes acceptedyear and that does not
#high p value, there is not a statistically significant difference between the two models, can
#drop the variable
anova(all.mod, all2.mod, test="LRT")


```

The C-statistic for the model without an accepted year is equal to 0.739 (see Table 3). Which shows that this model has a higher predictability capabilities compared to the model including the accepted year with a C-statistic = 0.683.

# Discussion
The key question at the heart of our study revolves around a fundamental question
healthcare: does a heart transplant prolong the life of patients? Logically, one would assume
that undergoing such a significant medical procedure, with all the risk involved and complexities,
should lead to an extended life of the individual. After all, why would patients willingly choose to
undergo heart transplants if they did not anticipate the promise of improved survival and a better
quality of life? To address this key question, we delved into the data provided by Stanford Heart
Transplant Study, utilizing the Cox Regression model to shed light on the matter.
Our analysis, as portrayed in Image 1, provides a compelling answer to the question. In
this graph, patients who received a heart transplant are represented in teal, while those who did
not receive a transplant are represented in red. The important observation is the remarkable
decrease in the rate of death among those who did not undergo a heart transplant over time.
The total sample size was 99 patients, 32 of them did not receive a transplant while 67 others
did receive the heart transplant. From the 32 who did not receive a transplant only 2 patients
managed to survive past 500 days. In stark contrast, among the 67 patients who did receive the
transplant, more than 22 of them survived beyond the 500-day mark. This visually striking
contrast suggests that heart transplantation does indeed have a significant impact on extending
the lives of patients in this context.
Image 2, on the other hand, offers a different perspective by categorizing patients based
on age groups; those under 40, those between 40 and 50, and those over 50 years old. The
calculated hazard ratio of 2.08 when moving from one age category to the next is a notable
finding. This ratio indicates that, all else being equal, patients in a higher age category are
approximately twice as likely to face mortality risks. This observation is irrespective of whether
they receive a heart transplant or not.
In a border sense, this implies that, within the confines of the criteria for inclusion in the
study, the need for a heart transplant–advancing from one age group to the next essentially
doubles the risk of mortality. This is entirely consistent with our understanding of the aging
process and its impact on an individual’s overall health. As one grows older, the body’s ability to
withstand the rigors of a medical condition that necessitates a heart transplant diminishes. The
data thus reinforces the notion that age itself is a critical factor in determining patients
outcomes.
In our group's pursuit of identifying the most robust and predictive model, we turned to
the concordance statistics, specifically the C-statistic, as a guiding metric for a model selection.
The C-statistic serves as a valuable tool for assessing the discriminatory power of predictive
models, aiding in our ability to distinguish their aptitude for accurately ranking risks and
predicting outcomes. Our findings are as follows:
$\textbf{Model 1: Excluding Wait Time}$
Our initial model was constructed without considering wait time as a variable. This
decision was rooted in the understanding that wait time data was only available for patients who
ultimately received a heart transplant, excluding those who did not have the opportunity to
undergo the procedure. The C-statistic for this model, as presented in Table 1, was calculated to
be 0.747. This initial C-statistic provided an encouraging start for our analysis.

\textbf{Model 2: Including Wait Time for Transplant Recipients}

The second model purpose was to assess the significance of wait time as a predictor of
outcomes but was excluded as an actual model due to the fact that it only applies to a subgroup
of patients who received a heart transplant. In this instance, the model excluded the control
group. The C-statistic for this model, focusing solely on transplant recipients, was 0.683, as
detailed in Table 2. However, it’s important to note that this model’s utility is limited due to its
omission of the control group, which restricts its generalizability and practicality.
Model 3: Variable Selection and ANOVA Table
Continuing our goal for the most informative model, we developed a third model. In this
iteration, we sought to explore the influence of the “accepted year” variable on predictive
capabilities. To make an informed decision regarding the necessity of including this variable, we
meticulously assessed its impact using an ANOVA table (as shown in Table 4). The analysis
provided us with a p-value of 0.1656, exceeding the common significance level of 0.05. As a
result, we concluded that “accepted year” could be excluded as a variable in the model.

\textbf{Model 3: Without Accepted Year}
With “accepted year” removed as a variable, we obtained a C-statistic of 0.739 for this
refined model, as indicated in Table. The improved C-statistic underscores the enhanced
predictive capabilities of this modelwhen compare to the previous iteration, which included
“accepted year” and had a C-statistic of 0.683

# Conclusion
In conclusion, our research provides strong support that heart transplantation has a
positive and significant impact on extending the lives of patients. The data was provided by
Stanford Heart Transplant Study. With the help of a survival analysis tool: Cox Regression
model. The Cox regression was best suited for this study due to the fact that it is suitable for
time-to-event outcomes. In this case, how much longer does a patient live if given a heart
transplant versus those who do not receive the transplant. In some cases patients post
transplant lived an extra 1-4 years. The best model used to support the data and build the
images was Model 3. Future research could look into which age group had best survivability
post transplant. This could provide healthcare workers valuable information regarding who
should get the heart transplant if there are patients in different age groups who are waiting for a
new heart and which would benefit the most and survive. As we know donated organs are not
easy to come by and making sure that an organ such as a heart should go to the best candidate
that would live longer with that new heart. Our findings contribute to a deeper understanding of
the complex interplay between heart transplantation, age and patient outcomes in the realm of
cardiac healthcare.

\textbf{Table 1: With variables: Accepted Year, Age, Prior, and Transplant}
```{r}
summary(all.mod)
```

\textbf{Table 2: With Variables Accepted Year, Age, Prior, and Wait Time}
```{r}
summary(treated.mod)
```

\textbf{Table 3: Model with Age, Prior, Transplant and Without Accepted Year}
```{r}
summary(all2.mod)
```

\textbf{Table 4: ANOVA Table Comparing Models 1 & 2}
```{r}
anova(all.mod, all2.mod, test="LRT")
```